Quantitative, Multispecies Monitoring at a Continental Scale

### 2025-02-12 Manuscript_plain.tex
Updates from Overleaf

\documentclass[9pt,twocolumn,twoside]{pnas-new}
% Use the lineno option to display guide line numbers if required.
\usepackage{soul} % Delete this line before submission XXX
\templatetype{pnasresearcharticle} % Choose template
% {pnasresearcharticle} = Template for a two-column research article
% {pnasmathematics} %= Template for a one-column mathematics article
% {pnasinvited} %= Template for a PNAS invited submission

@@

\author[a,1]{Gledis Guri}
\author[b]{Owen Liu}
\author[a]{Ryan P. Kelly}
\author[a]{Megan R. Shaffer}
\author[b]{Kim Parsons}
\author[a,b,c]{Ana Ram\'on-Laca}
\author[b]{Krista M. Nichols}
\author[a]{Pedro F. P. Brandão-Dias}
\author[b]{Abigail Wells}
\author[b]{Andrew Olaf Shelton}

\affil[a]{School of Marine and Environmental Affairs, University of Washington, Seattle, Washington, USA}
\affil[b]{Northwest Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA}
\affil[c]{Museo Nacional de Ciencias Naturales, Madrid, Spain}

% Please give the surname of the lead author for the running footer
\leadauthor{Guri}

% Please add a significance statement to explain the relevance of your work
\significancestatement{Detailed information about living resources is often legally required to support management decisions, but the cost and difficulty of acquiring this information increases dramatically at large spatial scales. We present a scaleable, quantitative method for mapping genetic trace material from species of management interest, and use it to illustrate three-dimensional patterns of environmental DNA (eDNA) from fish species along the West Coast of the United States. The same technique is applicable to environmental samples of any kind and any number of species, and is therefore likely to be broadly useful in natural-resources management and in the life sciences more generally.}

% Please include corresponding author, author contribution and author declaration information
\authorcontributions{
AOS, KMN, RPK, MS, and KP designed the research;
MRS, AR-L and AW did the laboratory work; 
PB, RPK, and GG did the bioinformatics;
GG, OL, RPK, and eAOS analyzed the data;
GG, OL, RPK, and AOS interpreted the results;
GG, OL, RPK, and AOS wrote the manuscript
\hl{[Please check that this is correct and add/edit as needed XXX]}
}
\authordeclaration{The authors declare no competing interest.}

\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: gguri@uw.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{eDNA $|$ Bayesian $|$ bio-monitoring $|$ quantification $|$ fish}

\begin{abstract}
Molecular data from environmental samples can reflect the abundance of species’ DNA, an index immediately relevant to natural-resources management at a broad geographic scale. These data commonly derive from assays designed and targeted for specific species (e.g., using qPCR or dPCR), or metabarcoding assays from more general PCR primers that amplify many species simultaneously. Multispecies analyses make efficient use of field samples and laboratory time, and speak to inherently multispecies questions of management and ecological research. However, unlike single-species techniques, metabarcoding by itself can reflect only the proportions of target-species DNA present, not their absolute quantities. Here, we combine qPCR and metabarcoding data derived from the same samples to map quantities of many fish species along the US West Coast in three dimensions, demonstrating a technique of practical importance for both management and ecology. We derive spatially explicit estimates of eDNA abundance for 12 common species of commercial and management importance and point the way to quantitative surveys of wild species using molecules alone. We find that species distribution maps derived from eDNA effectively capture species niches and spatial distributions. Notably, our analyses identified biodiversity hotspots that align with previously documented regions of ecological significance, such as the Columbia River plume and the Heceta Bank.
\end{abstract}

\dates{This manuscript was compiled on \today}
\doi{\url{www.pnas.org/cgi/doi/10.1073/pnas.XXXXXXXXXX}}

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}

\firstpage[4]{4}
Environmental DNA (eDNA, residual genetic material sampled from water, soil, or air) is an increasingly common tool for non-invasively sampling marine and aquatic ecological communities, valuable for estimating species occurrence \cite{veilleux2021}, biodiversity (i.e., Richness, \cite{muenzel2024}), and genetic structure \cite{andres2023a}. However, eDNA has not been widely used for quantitative abundance estimation, upon which many practical management and conservation applications rely (but see \cite{shelton2022,guri2024a,stoeckle2022a}). The use of eDNA for quantification is supported both conceptually -- where there is more of a species, more cells are shed in the environment, and thus there is more DNA -- and empirically (e.g. there are strong linkages between single-species eDNA measures of in aquaria \cite{jo2019b,ledger2024}, rivers and streams \cite{pont2024}, estuaries and nearshore habitats \cite{dibattista2022, baetscher2024}, and coastal oceans \cite{maes2023, shelton2019}). However, uses of quantitative eDNA observations in management applications have significantly lagged behind their use in occurrence and biodiversity applications.

@@

\section*{Results}

We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

@@

\begin{figure*}[tbhp]  % Use figure* for a column-wide figure
\centering
\includegraphics[width=16cm]{plots/1_Figure_1.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
\caption{Estimated species eDNA concentration (A) across 0, 50, 150, 300, and 500m depth samples and known depth distribution of those species from literature (B).}
\label{fig:fig1}
\end{figure*}

@@

\begin{figure}[tbhp]  % Use figure* for a column-wide figure
\centering
\includegraphics[width=8.5cm]{plots/2_Figure_2.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s rho correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig2}
\end{figure}

@@

\begin{figure*}[tbhp]  % Use figure* for a column-wide figure
\centering
\includegraphics[width=14cm]{plots/3_Figure_3.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
\caption{Spatial and vertical distribution of species richness (A) and environmental DNA (B) concentration across depths (0, 50, 150, 300, and 500 m). (B) Species richness represented as the cumulative number of species present (species with higher than 1.5 copies/L (see \href{SI_Appendix.pdf}{SI Appendix, Fig. S2} for defining presence) on smoothed estimated eDNA concentrations) and acumulated eDNA concentration representing the cumulative sum of eDNA (copies/L) across all species.}
\label{fig:fig3}
\end{figure*}

%\begin{figure*}[tbhp]  % Use figure* for a column-wide figure
%\centering
%\includegraphics[width=17cm]{plots/4_Figure_4.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
%\caption{Vertical (A) and latitudinal (B) distribution of eDNA concentrations grouped in ecological groups (surface south, surface north, midwater, and deepwater). Mean eDNA concentration of vertical distribution is shown with dots and 99\% upper quantile with ticks (A). Probability density functions (PDF) of estimated eDNA concentrations across latitudinal gradients (B; square root transformed of x-axis to enhance low concentration peaks). Key geographic features, such as the Columbia River and Heceta Bank, are marked to indicate alignment of species distributions.}
%\label{fig:fig4}
%\end{figure*}

\section*{Discussion}
We report species-specific DNA concentrations smoothed over ten degrees of latitude and across the continental shelf to 500m. We reveal distributional pattern for 12 species ranging from the iconic coastal pelagic species of the northeast Pacific to understudied-but-abundant deepwater species, and our results suggest clear pathways for the broader use of eDNA in both fisheries and ecosystem management.

@@

\matmethods{We summarize all methodological components and provide details for all sampling, laboratory, and statistical methods in the \href{SI_Appendix.pdf}{SI Appendix}.

@@

@@

\begin{figure*}[tbhp]  % Use figure* for a column-wide figure
\centering
\includegraphics[width=18cm]{plots/ppt_figures/DAG.pdf}  % Adjust width to 11.4 cm or 17.8 cm as per the template
\caption{Directed acyclic graph (DAG) of three models (orange, blue, and green) for estimating eDNA concentration (copies/L) for multiple (12) fish species. Model compartment indicated in the same color (2-5) are run jointly. Latent parameters, state variables, data and subscripts are indicated in gray. See \href{SI_Appendix.pdf}{SI Appendix} for full model breakdown.}
\label{fig:DAG}
\end{figure*}

@@

}

\showmatmethods{} % Display the Materials and Methods section

\acknow{We thank Linda Park for her foresight and perseverance, Katherine Maslenikov (University of Washington Burke Museum Ichthyology Collection) for providing tissue samples for the mock community analysis, and Olivia Scott for validation and Sanger sequencing of the mock communities. This material is based upon research supported by the Office of Naval Research under Award Number (N00014-22-1-2719). The authors gratefully acknowledge the survey expertise and support of NOAA's Pacific hake survey team and the personnel of the NOAA Ship Bell M. Shimada for supporting the collection of eDNA samples on the west coast hake survey 2019 - 2023. Any use of trade, firm, or product names is for descriptive purposes only and does not imply endorsement by the U.S. Government.}

\showacknow{} % Display the acknowledgments section

\bibsplit[9]
%Use \bibsplit to split the references from the body of the text. Value "[2]" represents the number of reference in the left column (Note: Please avoid single column figures & tables on this page.)
% Bibliography
\bibliography{pnas-sample}

>>>>

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{soul}
\bibliographystyle{plain}

\author{Gledis Guri$^1$\textbf{*} \and
Owen Liu$^2$ \and
Ryan P. Kelly$^1$ \and
Megan R. Shaffer$^1$\and
Kim Parsons$^2$ \and
Ana Ram\'on-Laca$^{1,2,3}$ \and
Krista M. Nichols$^2$ \and
Pedro F. P. Brandão-Dias$^1$ \and
Abigail Wells$^2$ \and
Andrew Olaf Shelton$^2$ \and}

\date{\today}

@@

\maketitle

\section*{}

\begin{center}
\begin{tabular}{ll}
1 & School of Marine and Environmental Affairs, University of Washington, Seattle, Washington, USA \\
2 & Northwest Fisheries Science Center, NOAA Fisheries, Seattle, Washington, USA \\
3 & Museo Nacional de Ciencias Naturales, Madrid, Spain \\
\hline
\textbf{*} & \textbf{Gledis Guri}, lead and corresponding author \textbf{gguri@uw.edu}
\end{tabular}
\end{center}

\section*{Significance statement}
Detailed information about living resources is often legally required to support management decisions, but the cost and difficulty of acquiring this information increases dramatically at large spatial scales. We present a scaleable, quantitative method for mapping genetic trace material from species of management interest, and use it to illustrate three-dimensional patterns of environmental DNA (eDNA) from fish species along the West Coast of the United States. The same technique is applicable to environmental samples of any kind and any number of species, and is therefore likely to be broadly useful in natural-resources management and in the life sciences more generally.

\section*{Author Contributions}
AOS, KMN, RPK, MS, and KP designed the research;\\
MRS, AR-L and AW did the laboratory work; \\
PB, RPK, and GG did the bioinformatics;\\
GG, OL, RPK, and eAOS analyzed the data;\\
GG, OL, RPK, and AOS interpreted the results;\\
GG, OL, RPK, and AOS wrote the manuscript

[Please check that this is correct and add/edit as needed XXX]

\section*{Author Declaration}
The authors declare no competing interest.

\section*{Keywords}
eDNA $|$ Bayesian $|$ bio-monitoring $|$ quantification $|$ fish

\section*{Abstract}
Molecular data from environmental samples can reflect the abundance of species’ DNA, an index immediately relevant to natural-resources management at a broad geographic scale. These data commonly derive from assays designed and targeted for specific species (e.g., using qPCR or dPCR), or metabarcoding assays from more general PCR primers that amplify many species simultaneously. Multispecies analyses make efficient use of field samples and laboratory time, and speak to inherently multispecies questions of management and ecological research. However, unlike single-species techniques, metabarcoding by itself can reflect only the proportions of target-species DNA present, not their absolute quantities. Here, we combine qPCR and metabarcoding data derived from the same samples to map quantities of many fish species along the US West Coast in three dimensions, demonstrating a technique of practical importance for both management and ecology. We derive spatially explicit estimates of eDNA abundance for 12 common species of commercial and management importance and point the way to quantitative surveys of wild species using molecules alone. We find that species distribution maps derived from eDNA effectively capture species niches and spatial distributions. Notably, our analyses identified biodiversity hotspots that align with previously documented regions of ecological significance, such as the Columbia River plume and the Heceta Bank.

\section{Introduction}
Environmental DNA (eDNA, residual genetic material sampled from water, soil, or air) is an increasingly common tool for non-invasively sampling marine and aquatic ecological communities, valuable for estimating species occurrence \cite{veilleux2021}, biodiversity (i.e., Richness, \cite{muenzel2024}), and genetic structure \cite{andres2023a}. However, eDNA has not been widely used for quantitative abundance estimation, upon which many practical management and conservation applications rely (but see \cite{shelton2022,guri2024a,stoeckle2022a}). The use of eDNA for quantification is supported both conceptually -- where there is more of a species, more cells are shed in the environment, and thus there is more DNA -- and empirically (e.g. there are strong linkages between single-species eDNA measures of in aquaria \cite{jo2019b,ledger2024}, rivers and streams \cite{pont2024}, estuaries and nearshore habitats \cite{dibattista2022, baetscher2024}, and coastal oceans \cite{maes2023, shelton2019}). However, uses of quantitative eDNA observations in management applications have significantly lagged behind their use in occurrence and biodiversity applications.

@@

\section{Results}
We find spatial patterns of DNA concentrations reflect the distributions of commercially important coastal-pelagic fish species in three dimensions and across ten degrees of latitude in a single year of sampling in 2019 (Fig. \ref{fig:fig1}: Northern Anchovy, Pacific Sardine; \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}, Pacific Herring and Jack Mackerel). For other high-biomass species otherwise lacking well-described distributions, we provide the first species distribution models (Fig. \ref{fig:fig1}, \href{SI_Appendix.pdf}{SI Appendix, Fig. S1}). We find that the estimated eDNA concentrations are correlated among species with similar traits and habitat preferences (Fig. \ref{fig:fig2}). Emergent patterns of species richness and estimated eDNA aggregation correspond with known ecological drivers, evidence that eDNA traces can reliably capture ecologically meaningful patterns (Fig. \ref{fig:fig3}). Particularly for forage fish species – which are critical high-biomass links in marine food webs, but difficult to survey with traditional methods \cite{Pikitch2012} – molecular methods offer a tractable and quantitative index of abundance over large spatial scales.

@@

\begin{figure*}[tbhp]
\centering
\includegraphics[width=16cm]{plots/1_Figure_1.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
\caption{Estimated species eDNA concentration (A) across 0, 50, 150, 300, and 500m depth samples and known depth distribution of those species from literature (B).}
\label{fig:fig1}
\end{figure*}

@@

\begin{figure}[tbhp] 
\centering
\includegraphics[width=8.5cm]{plots/2_Figure_2.jpg}  
\caption{Correlation plot describing species co-abundance and co-occurance. Spearman’s rho correlation coefficient are expressed in numbers and colors ranging from -1 (red) negatively correlated to 1 (blue) positively correlated. Species are ordered in known ecological groups (surface, midwater, and deepwater) to visually indicate the gradual change of species co-occurance.}
\label{fig:fig2}
\end{figure}

@@

\begin{figure*}[tbhp]  
\centering
\includegraphics[width=13cm]{plots/3_Figure_3.jpg}  
\caption{Spatial and vertical distribution of species richness (A) and environmental DNA (B) concentration across depths (0, 50, 150, 300, and 500 m). (B) Species richness represented as the cumulative number of species present (species with higher than 1.5 copies/L (see \href{SI_Appendix.pdf}{SI Appendix, Fig. S2} for defining presence) on smoothed estimated eDNA concentrations) and acumulated eDNA concentration representing the cumulative sum of eDNA (copies/L) across all species.}
\label{fig:fig3}
\end{figure*}

%\begin{figure*}[tbhp]  
%\centering
%\includegraphics[width=17cm]{plots/4_Figure_4.jpg}  % Adjust width to 11.4 cm or 17.8 cm as per the template
%\caption{Vertical (A) and latitudinal (B) distribution of eDNA concentrations grouped in ecological groups (surface south, surface north, midwater, and deepwater). Mean eDNA concentration of vertical distribution is shown with dots and 99\% upper quantile with ticks (A). Probability density functions (PDF) of estimated eDNA concentrations across latitudinal gradients (B; square root transformed of x-axis to enhance low concentration peaks). Key geographic features, such as the Columbia River and Heceta Bank, are marked to indicate alignment of species distributions.}
%\label{fig:fig4}
%\end{figure*}

\section{Discussion}
We report species-specific DNA concentrations smoothed over ten degrees of latitude and across the continental shelf to 500m. We reveal distributional pattern for 12 species ranging from the iconic coastal pelagic species of the northeast Pacific to understudied-but-abundant deepwater species, and our results suggest clear pathways for the broader use of eDNA in both fisheries and ecosystem management.

@@

\section{Methods}
We summarize all methodological components and provide details for all sampling, laboratory, and statistical methods in the \href{SI_Appendix.pdf}{SI Appendix}.

@@

\begin{figure*}[tbhp] 
\centering
\includegraphics[width=16cm]{plots/ppt_figures/DAG.pdf} 
\caption{Directed acyclic graph (DAG) of three models (orange, blue, and green) for estimating eDNA concentration (copies/L) for multiple (12) fish species. Model compartment indicated in the same color (2-5) are run jointly. Latent parameters, state variables, data and subscripts are indicated in gray. See \href{SI_Appendix.pdf}{SI Appendix} for full model breakdown.}
\label{fig:DAG}
\end{figure*}

@@

@@

\bibliography{pnas-sample.bib}
